---
title: "MESA-MIND-Longitudinal-Proteomics-cSVD"
author: "LekkiWood (LekkiWood@Gmail.com)"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
crossref:
  tbl-prefix: "Table"
  fig-prefix: "Figure"
execute:
  echo: false
  warning: false
  message: false
  # ensure chunks run from the project root (where _targets/ lives)
  dir: project
pdf-engine: lualatex
editor: visual
---

```{r}
#| label: load-packages


## Load targets
if (!requireNamespace("targets", quietly = TRUE)) {
  stop("The {targets} package is not installed/available in this render session.")
}

## Load knitr
if (!requireNamespace("knitr", quietly = TRUE)) {
  stop("The {knitr} package is not installed/available in this render session.")
}

## Load quarto
if (!requireNamespace("quarto", quietly = TRUE)) {
  stop("The {quarto} package is not installed/available in this render session.")
}

## Load gtsummary
if (!requireNamespace("gtsummary", quietly = TRUE)) {
  stop("The {gtsummary} package is not installed/available in this render session.")
}


```

\newpage

# Important Notes - read me first

-   Always check that you have the most recent version of this document, which - unless I am sending you unfinalized work - is available [here](https://github.com/LekkiWood/MESA-MIND-Longitudinal-Proteomics-cSVD/blob/main/MESA-MIND-Longitudinal-Proteomics-cSVD.pdf).

    -   An easy check for version control is to make sure this date: `r Sys.Date()`. is the same as on the GitHub file [here](https://github.com/LekkiWood/MESA-MIND-Longitudinal-Proteomics-cSVD/blob/main/MESA-MIND-Longitudinal-Proteomics-cSVD.pdf).

-   The code for this is analysis available in the same repository ([targets master file here](https://github.com/LekkiWood/MESA-MIND-Longitudinal-Proteomics-cSVD/blob/main/_targets.R) and [individual functions here](https://github.com/LekkiWood/MESA-MIND-Longitudinal-Proteomics-cSVD/tree/main/R))

# Summary

# Step 1: Cleaning and Formatting Proteins

```{r}

#| label: load-filename-info

#General QC info
QC_info <- targets::tar_read(build_proteins_QC)

#Final N
finalclean_N_by_exam <- targets::tar_read(Proteins_clean_N)
```

## Input file names

-   A table of protein abundances: `{r} print(unlist(strsplit(QC_info$filenames$raw_proteins_file_name, split="/"))[8])`
-   Sample information to link TOPMed IDs to unique MESA SHARe ID and exam combinations: `{r} print(unlist(strsplit(QC_info$filenames$raw_protein_info_file_name, split="/"))[8])`
-   Keys to link Olink IDs to names compounds: `{r} print(unlist(strsplit(QC_info$filenames$ raw_protein_keys_file_name, split="/"))[8])`
-   A file to bridge SHARe ids (sidno) with MESA IDs (idno) `{r} print(unlist(strsplit(QC_info$filenames$raw_bridgingfile_file_name, split="/"))[7])`

## Raw file info

-   The raw protein abundance file contained information on N=`{r} print(QC_info$raw_inputs$raw_protein_N_all)` protein assays, including those used for QC.

-   When removing assays for QC, the raw protein abundance file contained information on N=`{r} print(QC_info$raw_inputs$raw_protein_N_proteins[1,1])` proteins.

-   The protein abundance file contained information on N=`{r} print(QC_info$raw_inputs$raw_N_all)` sample IDs (i.e., unique participant/exam combinations), including bridging samples.

-   After removing QC samples (including bridging, controls, and one duplicate) the protein abundance file contained information on N=`{r} print(QC_info$raw_inputs$raw_N_nobridge)` sample IDs (i.e., unique participant/exam combinations).

### Formatting

-   Bridging (and other QC) samples were removed.

-   Protein assays used for QC were removed.

-   Proteins that should be excluded due to QC warnings (variable "QC_warning" set to "EXCLUDED") were removed, even though these do not have NPX values.

-   Data were put into wide format, with "SampleID" as the unique ID, "OlinkID" forming the variable names (protein identifiers), and values taken from the "NPX" column.

    -   In wide format, the file contained information on N=`{r} print(QC_info$formatted_protein_file_info$unique_sampleids)` unique sample IDs.

    -   In wide format, the file contained information on N=`{r} print(QC_info$formatted_protein_file_info$duplicated_sampleids)` duplicated sample IDs. ^1^

-   SHARe IDs, and subsequently MESA IDs, were merged into the file with exam information.

-   At this point, the range of unique SHARe ID by exam combinations was N=`{r} range(QC_info$initial_dupes$Freq)[1]` - `{r} range(QC_info$initial_dupes$Freq)[2]`. This indicates no sample ID were duplicated in the assays.

-   The formatted protein file was used to calculate the coefficient of variation (CV) using the formula: CV = sqrt(2`^`(sigma`^`2)-1).

-   A variable called "Retain" was created to indicate whether each protein was (1) unique (i.e., included on only one panel); (2) duplicated, and across all panels had the lowest CV; or (3) duplicated, and across all panels did not have the lowest CV.

-   A final table of protein abundances, with additional variables for SHARe ID, MESA ID, Exam, TOPMed ID and Batch, was created after the steps above, with proteins duplicated across more than one panel cleaned such that only the one with the lowest CV is retained. This file was used in the analysis

-   This file included data on the following numbers of participants, stratified by exam:

```{r}

knitr::kable(finalclean_N_by_exam, format="latex") #print(knitr::kable(finalclean_N_by_exam))
```

\newpage

# Step 2: Format Phenotypes

```{r}

#| label: read-in-trait-info

Traits_QC_info <- targets::tar_read(traits_QC_info)

Traits_db <- targets::tar_read(traits_db)
```

## Input files

-   Covariates from E1: `{r} print(unlist(strsplit(Traits_QC_info$filenames$E1_covs_filename, split="/"))[7])`

-   Covariates from E5: `{r} print(unlist(strsplit(Traits_QC_info$filenames$E5_covs_filename, split="/"))[7])`

-   Covariates from E6: `{r} print(unlist(strsplit(Traits_QC_info$filenames$E6_covs_filename, split="/"))[7])`

-   Afib info: `{r} print(unlist(strsplit(Traits_QC_info$filenames$afib_filename, split="/"))[7])`

-   ApoE info: `{r} print(unlist(strsplit(Traits_QC_info$filenames$apoe_filename, split="/"))[7])`

-   Incident CVD: `{r} print(unlist(strsplit(Traits_QC_info$filenames$cvd_filename, split="/"))[7])`

-   Microbleeds: `{r} print(unlist(strsplit(Traits_QC_info$filenames$mb_filename, split="/"))[7])`

-   Perivascular spaces: `{r} print(unlist(strsplit(Traits_QC_info$filenames$evps_filename, split="/"))[7])`

-   White matter hyperintensities: `{r} print(unlist(strsplit(Traits_QC_info$filenames$wmh_filename, split="/"))[7])`

-   Intracranial volumes: `{r} print(unlist(strsplit(Traits_QC_info$filenames$icv_filename, split="/"))[7])`

-   Fractional anisotropy: `{r} print(unlist(strsplit(Traits_QC_info$filenames$wmfa_filename, split="/"))[7])`

-   White matter hyperintensities: `{r} print(unlist(strsplit(Traits_QC_info$filenames$wmh_filename, split="/"))[7])`

## Formatting

### Outcomes:

-   Microbleeds were coded as 0/1, where 0= no microbleeds (value: 0) and 1 = presence of microbleeds (all non-zero values except missing). Then, those images with a low image quality (value = 4; N=`{r} length(Traits_db$qsm_swi_image_quality[Traits_db$qsm_swi_image_quality == 4 & !is.na(Traits_db$qsm_swi_image_quality)])` ) were recoded to missing.

```{r}
#| label: raw-MB-N-quality

knitr::kable(table(Traits_db$mb_present), format="latex")
```

-   Perivascular spaces (variable: epvs_wholebrain_vol) were recoded to missing where the variable 'pvs_exclude' was coded as 1 (N=`{r} length(Traits_db[Traits_db$pvs_exclude ==1 & !is.na(Traits_db$pvs_exclude),]$pvs_exclude)`).

-   White matter hyperintensities (variable: wm_wmh) were divided by 1000 to convert to ml (following Rizwan's code), and those where the variable wmh_exclude had a code of 1 were set to missing.

-   Fractional anisotropy (variable: wmfa) was coded to missing where the variable fa_exclude had a value of 1.

## Covariates

#### True time invariant covariates

-   Race/ethnicity, gender, and highest education level were all taken from exam 1 data. ApoE information was taken from its own dataset (above).

    -   ApoE was coded 0/1/2 where 0= no e4 isoform (codes 22, 23, 33), 1 = e4 isoform (24, 34, 44), and 2 = no isoform data. The ApoE variable was formatted as a factor.

    -   Gender was coded such that female = 0 and male = 1.

    -   Education was recoded 0/1, such that 0 = less than high school (codes: 0: NO SCHOOLING / 1: GRADES 1-8 / 2: GRADES: 9-11) and 1= high school or more (all other codes, excluding missing).

    -   Race/ethnicity was recoded retaining the original MESA coded whereby 1=White American; 2= Chinese American, 3=Black, African-American, and 4 = Hispanic. Race/ethnicity was coded as a factor variable.

#### Pseudo-time invariant covariates

-   Although some variables are technically time invariant, where they were included due to their effects on MRI data, since MRI data are only measured at one exam for this analysis (exam 6), these covariates were always taken from exam 6.

-   These 'pseudo time invariant covariates' were: atrial fibrillation, myocardial infarction, congestive heart failure, LDL, systolic blood pressure, hypertension medication, and site (since site seems to affect MRI more than proteins??).

    -   Afib, MI, and CHF were coded 0/1, such that 0= no diagnosis and 1= afib diagnosis. Missing data was left as missing (this is different to Rizwan who coded missing data as no diagnosis).

    -   Site was arbitrarily coded as 0=Wake forest, 1= Columbia, 2=Johns Hopkins, 3=University of Minnesota, 4=Northwestern, 5=UCLA

    #### Time varying covariates

-   The following covariates were taken from the exam when the proteins were used, as these were seen to affect proteins more in the short term than they affect MRI (?): kidney function (egfr), BMI , cigarette smoking (never/former/current; coded as ordinal), diabetes status.

    -   Smoking was harmonized and coded such that 0 = never smoker, 1= past smoker, 2=current smoker.

    -   Diabetes has harmonized and coded such that 0= no diabetes (including impaired fasting glucose), and 1 = diabetes (treated and untreated).

-   The following covariates were taken from exam 6: age (age6c), kidney function (egfr; cepgfr6c), BMI (bmi6c), systolic blood pressure (sbp6c), LDL (ldl6), site(site6c), the use of hypertension medication (htnmed6c; coded 0= no, and 1= yes), cigarette smoking (cig6c), diabetes status (dm036t)

# Step 3: Sample descriptives

```{r}

#| label: info-for-descrptives

mind_ids <- targets::tar_read(mind_ids)
protein_and_MIND_ids <- targets::tar_read(protein_and_MIND_ids)
protein_and_MIND_and_cov_ids <- targets::tar_read(protein_and_MIND_and_cov_ids)
```

-   There were N=`{r} print(length(unique(mind_ids$idno)))` MESA participants with at least one MRI outcome after the exclusions above.

-   Of those with MRI data, N=`{r} print(length(unique(protein_and_MIND_ids$idno)))` participants had protein data, equating to N=`{r} length(unique(protein_and_MIND_ids[which(protein_and_MIND_ids$Exam==1),]$idno))` at exam 1, N=`{r} length(unique(protein_and_MIND_ids[which(protein_and_MIND_ids$Exam==5),]$idno))` at exam 5, and N=`{r} length(unique(protein_and_MIND_ids[which(protein_and_MIND_ids$Exam==6),]$idno))` at exam 6.

-   Of those with protein data and at least one MRI outcome, N=`{r} print(length(unique(protein_and_MIND_and_cov_ids$idno)))` participants had no missing covariate data and so were included in this analysis, equating to N=`{r} length(unique(protein_and_MIND_and_cov_ids[which(protein_and_MIND_and_cov_ids$Exam==1),]$idno))` at exam 1, N=`{r} length(unique(protein_and_MIND_and_cov_ids[which(protein_and_MIND_and_cov_ids$Exam==5),]$idno))` at exam 5, and N=`{r} length(unique(protein_and_MIND_and_cov_ids[which(protein_and_MIND_and_cov_ids$Exam==6),]$idno))` at exam 6.

::: landscape
```{r}

#| label: table of descriptives

bold_labels <- c(
  "Age (y)", "Gender", "Field Center", "Race or ethnicity", "Highest education level",
  "BMI (kg/m^2^)", "LDL levels", "systolic blood pressure", "Diabetes status",
  "Takes hypertentsion medicine", "Atrial fibrillation", "Myocardial Infarction",
  "Coronary Heart Failure", "ApoeE information", "Smoking status",
  "Kidney function (egfr)", "Intracranial volume", "Fractional anisotropy",
  "White matter hyperintensities", "Enlarged perivascular spaces",
  "Presence of microbleeds?"
)



table1 <- Traits_db |> 
  dplyr::mutate(
        gender_desc = factor(gender,
                       levels = c("0", "1"),
                       labels = c("Female", "Male")),
        Exam_desc = factor(Exam,
                       levels = c("1", "5", "6"),
                       labels = c("Exam 1", "Exam 5", "Exam6")),
        race_desc = factor(race,
                       levels = c("1", "2", "3", "4"),
                       labels = c("Non-Hispanic White", "Chinese American", "Black/African-American", "Hispanic")),
        mb_present_desc = factor(mb_present,
                       levels = c("0", "1"),
                       labels = c("No", "Yes")),
                       smoking_desc = factor(smoking,
                       levels = c("0", "1", "2"),
                       labels = c("Never", "Former", "Current")),
        diabetes_desc = factor(diabetes,
                       levels = c("0", "1"),
                       labels = c("Normoglycemia/IFG", "Diabetes (treated or untreated)")),
        htnmeds_desc = factor(htnmeds,
                       levels = c("0", "1"),
                       labels = c("No", "Yes")),
        edu_desc = factor(edu,
                       levels = c("0", "1"),
                       labels = c("Up to and including high school", "More than high school")),
        E4_desc = factor(E4,
                       levels = c("0", "1", "2"),
                       labels = c("No E4 isoform", "E4 isoform", "No ApoE data")),
        site_desc = factor(site,
                       levels = c("1", "2", "3", "4", "5", "6"),
                       labels = c("Wake Forest", "Columbia","Johns Hopkins", "Minnesota", "Northwestern", "UCLA")),
         AFprevalent_desc = factor(AFprevalent,
                       levels = c("0", "1"),
                       labels = c("No", "Yes")),
        MIprevalent_desc = factor(MIprevalent,
                       levels = c("0", "1"),
                       labels = c("No", "Yes")),
         CHFprevalent_desc = factor(CHFprevalent,
                       levels = c("0", "1"),
                       labels = c("No", "Yes")),
        smoking_desc = factor(smoking,
                       levels = c("0", "1", "2"),
                       labels = c("Never", "Former", "Current"))
      ) |>
gtsummary::tbl_summary(
    include = c(age, gender_desc, site_desc, edu_desc, race_desc, BMI, smoking_desc, ldl, sbp, diabetes_desc, htnmeds_desc, AFprevalent_desc, MIprevalent_desc, CHFprevalent_desc, E4_desc, egfr, icv, fa, wmh, epvs, mb_present_desc),
    by = Exam,
    statistic = list(
      gtsummary::all_continuous() ~ "{mean} ({sd})",
      gtsummary::all_categorical() ~ "{n} / {N} ({p}%)"
    ),# split table by group
    missing = "no", # don't list missing data separately
    digits = gtsummary::all_continuous() ~ 2, 
   ##add in labels
     label = list(age = "Age (y)",
                 gender_desc = "Gender",
                 site_desc = "Field Center", 
                 race_desc = "Race or ethnicity",
                 edu_desc = "Highest education level",
                 BMI = "BMI (kg/m^2^)",
                 ldl = "LDL levels",
                 sbp = "systolic blood pressure",
                 diabetes_desc = "Diabetes status",
                 htnmeds_desc = "Takes hypertentsion medicine",
                 AFprevalent_desc = "Atrial fibrillation",
                 MIprevalent_desc = "Myocardial Infarction",
                 CHFprevalent_desc = "Coronary Heart Failure",
                 E4_desc = "ApoeE information",
                 smoking_desc = "Smoking status",
                 egfr = "Kidney function (egfr)",
                 icv = "Intracranial volume",
                 fa ="Fractional anisotropy",
                 wmh="White matter hyperintensities",
                 epvs="Enlarged perivascular spaces",
                 mb_present_desc = "Presence of microbleeds?"),
   type = list(age ~ "continuous",
                 gender_desc ~ "categorical",
                 site_desc ~ "categorical", 
                 race_desc ~ "categorical",
                 edu_desc ~ "categorical",
                 BMI ~ "continuous",
                 ldl ~ "continuous",
                 sbp ~ "continuous",
                 diabetes_desc ~ "categorical",
                 htnmeds_desc ~ "categorical",
                 AFprevalent_desc ~ "categorical",
                 MIprevalent_desc ~ "categorical",
                 CHFprevalent_desc ~ "categorical",
                 E4_desc ~ "categorical",
                 smoking_desc ~ "categorical",
                 egfr ~ "continuous",
                 icv ~ "continuous",
                 fa ~ "continuous",
                 wmh ~ "continuous",
                 epvs ~ "continuous",
                 mb_present_desc ~ "categorical"
    ) 
) |>
  gtsummary::modify_spanning_header(
    gtsummary::all_stat_cols() ~ "Exam"
  ) |>
  gtsummary::as_gt() |>
  gt::tab_options(page.orientation = "landscape") |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_body(
      columns = "label",
      rows = label %in% bold_labels
    )
    ) |>
# Bold *any* spanning headers (includes your "Exam" spanner)
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_spanners()
  ) |>
  gt::tab_options(latex.use_longtable = TRUE)

table1


```
:::

# Notes

## Footnotes

^1^ This is a reproducible file for many runs, containing many data checks. Values of 0 or NULL are expected, and just indicate no problem with the data.

## Session Info

For reproducibility

```{r}

#| label: session-info

# Prefer the detailed sessioninfo output; otherwise fall back to base.
if (requireNamespace("sessioninfo", quietly = TRUE)) {
  sessioninfo::session_info()
} else {
  sessionInfo()
}
```
